---
title: "Homework 3 – Problem 2: Neural Bayes Estimators"
author: "Team Members: [Names Here]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(2026)
```

# Background

We consider the exponentially tilted distribution with density:

$$
f(x;\theta) = \sqrt{\frac{1}{4\pi}} \cdot \frac{\exp\!\left(-\frac{1}{4x} - \theta x\right)}{x^{3/2}\,\exp\!\left(-\sqrt{\theta}\right)}, \quad x > 0,\; \theta > 0.
$$

Rearranging, we can write the log-density as:

$$
\log f(x;\theta) = -\tfrac{1}{2}\log(4\pi) - \tfrac{3}{2}\log x - \frac{1}{4x} - \theta x + \sqrt{\theta}.
$$

This is an exponentially tilted inverse-Gaussian form. When $\theta = 0$ the
density reduces to a Lévy($0, 1/2$) distribution (one-sided stable with index
1/2). The tilting parameter $\theta$ controls how quickly the right tail decays.

Each test dataset consists of $k = 50$ i.i.d. draws from this density under a
common $\theta$.

## Load the test data

```{r load-data}
# ------------------------------------------------------------------
# IMPORTANT: place test_data.RData in your working directory first
# ------------------------------------------------------------------
load("test_data.RData")          # should create an object (list of 1000 datasets)

# Inspect the structure – adapt the object name if different
# Common names the instructor might use: test_data, Z_test, datasets
# Uncomment and adjust as needed:
# str(test_data)
# length(test_data)

# We assume the loaded object is called Z_test (a list of 1000 vectors, each length 50).
# If the object has a different name, rename it here:
# Z_test <- test_data          # <-- uncomment / adjust if needed

cat("Number of test datasets:", length(Z_test), "\n")
cat("Size of first dataset:", length(Z_test[[1]]), "\n")
head(Z_test[[1]])
```

# Part 1 – Maximum Likelihood Estimation (30 pts)

## Log-likelihood

For a single dataset $\mathbf{x} = (x_1,\dots,x_k)$ the log-likelihood is

$$
\ell(\theta) = \sum_{i=1}^{k}\log f(x_i;\theta)
  = C(\mathbf{x}) - \theta \sum_i x_i + k\sqrt{\theta},
$$

where $C(\mathbf{x}) = -\frac{k}{2}\log(4\pi) - \frac{3}{2}\sum_i\log x_i - \frac{1}{4}\sum_i \frac{1}{x_i}$
does not depend on $\theta$.

Setting the derivative to zero:

$$
\ell'(\theta) = -\sum_i x_i + \frac{k}{2\sqrt{\theta}} = 0
  \quad\Longrightarrow\quad
  \hat\theta_{\text{MLE}} = \left(\frac{k}{2\sum_i x_i}\right)^2
  = \frac{1}{(2\bar x)^2}.
$$

Because the second derivative $\ell''(\theta) = -\frac{k}{4}\theta^{-3/2} < 0$, this is
indeed a maximum.

**Special case $\theta = 0$**: The MLE formula always gives $\hat\theta > 0$, but
for datasets generated under $\theta = 0$ or very small $\theta$, the data will have
very heavy tails and large $\bar x$, pushing $\hat\theta$ toward zero naturally.

## Implementation

```{r mle}
# Closed-form MLE
mle_theta <- function(x) {
  xbar <- mean(x)
  (1 / (2 * xbar))^2
}

# Compute MLE for all 1000 test datasets
theta_mle <- sapply(Z_test, mle_theta)

cat("Summary of MLE estimates:\n")
summary(theta_mle)
```

```{r mle-hist, fig.width=7, fig.height=4}
hist(theta_mle, breaks = 50, col = "steelblue", border = "white",
     main = "Distribution of MLE estimates across 1000 test datasets",
     xlab = expression(hat(theta)[MLE]))
```

# Part 2 – Neural Bayes Estimation (45 pts)

We use the **NeuralEstimators** R package (which interfaces with Julia)
to build an amortised neural Bayes estimator. The idea: train a neural
network on many simulated datasets so that it learns to map a dataset
of size $k = 50$ to a point estimate of $\theta$.

## Prerequisites

You need Julia installed plus the Julia packages `NeuralEstimators` and
`Flux`. For GPU support, also install `CUDA` in Julia.

```{r nbe-setup, eval=FALSE}
# ----- One-time Julia setup (run in Julia REPL, not R) -----
# using Pkg
# Pkg.add("NeuralEstimators")
# Pkg.add("Flux")
# Pkg.add("CUDA")        # for GPU acceleration (optional)
# Pkg.add("Distributions")

# ----- R packages -----
# install.packages("NeuralEstimators")
# install.packages("JuliaConnectoR")
```

```{r nbe-libraries, eval=FALSE}
library(NeuralEstimators)
library(JuliaConnectoR)
juliaEval("using NeuralEstimators, Flux, Distributions")

# Check GPU availability
juliaEval("using CUDA; CUDA.functional()")
```

## Simulating training data

The homework says $\theta$ values in the test set are concentrated near 0
(including $\theta=0$ exactly) but spread across $[0, 2]$. Our training
prior should cover this range and put extra mass near the boundary.

**Training prior for $\theta$**: We use a mixture —
70 % from $\text{Exponential}(\text{rate}=3)$ (concentrated near 0) and
30 % from $\text{Uniform}(0, 2.5)$ (covering the full range, slightly wider
than [0,2] to provide a buffer).

**Simulator**: To sample $X \sim f(\cdot;\theta)$ we use the fact that
$f(x;\theta)$ is an inverse-Gaussian distribution.
Specifically, matching the standard parameterisation
$\text{IG}(\mu, \lambda)$ with $\mu = 1/(2\sqrt{\theta})$ and $\lambda = 1/2$
gives us the desired density (for $\theta > 0$).
For $\theta = 0$, the distribution becomes the Lévy distribution; we can
simulate it as $X = 1/(4Z^2)$ where $Z \sim N(0,1)$.

```{r nbe-sampler-simulator, eval=FALSE}
# ---- Prior sampler: returns a 1 x K matrix of theta values ----
sampler <- function(K) {
  theta_vals <- numeric(K)
  for (i in 1:K) {
    if (runif(1) < 0.10) {
      theta_vals[i] <- 0                      # point mass at 0
    } else if (runif(1) < 0.667) {
      theta_vals[i] <- rexp(1, rate = 3)      # concentrated near 0
    } else {
      theta_vals[i] <- runif(1, 0, 2.5)       # spread across range
    }
  }
  matrix(theta_vals, nrow = 1)
}

# ---- Inverse-Gaussian sampler (Michael/Schucany/Haas method) ----
rinvgauss <- function(n, mu, lambda) {
  # Generates n draws from IG(mu, lambda)
  nu <- rnorm(n)
  y  <- nu^2
  x  <- mu + (mu^2 * y) / (2 * lambda) -
    (mu / (2 * lambda)) * sqrt(4 * mu * lambda * y + mu^2 * y^2)
  z  <- runif(n)
  ifelse(z <= mu / (mu + x), x, mu^2 / x)
}

# ---- Data simulator: returns a list of 1 x m matrices ----
simulator <- function(theta_set, m) {
  # theta_set is a 1 x K matrix
  apply(theta_set, 2, function(theta) {
    if (theta < 1e-12) {
      # theta ≈ 0: Lévy distribution  X = 1/(4*Z^2)
      z <- rnorm(m)
      x <- 1 / (4 * z^2)
    } else {
      mu_ig     <- 1 / (2 * sqrt(theta))
      lambda_ig <- 0.5
      x <- rinvgauss(m, mu_ig, lambda_ig)
    }
    matrix(x, nrow = 1)
  }, simplify = FALSE)
}
```

## Neural network architecture

We use a **DeepSets** architecture — ideal for exchangeable (i.i.d.) data.
The inner network $\psi$ processes each observation independently, and
after aggregation (mean pooling), the outer network $\phi$ maps the
summary to a parameter estimate.

Since $\theta \ge 0$ we use a softplus output to ensure non-negativity.

```{r nbe-architecture, eval=FALSE}
estimator <- juliaEval('
  d = 1    # dimension of each observation
  p = 1    # number of parameters to estimate (theta)
  w = 128  # hidden layer width

  psi = Chain(Dense(d, w, relu), Dense(w, w, relu), Dense(w, w, relu))
  phi = Chain(Dense(w, w, relu), Dense(w, w, relu), Dense(w, p, softplus))

  deepset = DeepSet(psi, phi)
  PointEstimator(deepset)
')
```

## Loss function

We use the default **mean-absolute-error (MAE)** loss, which means
the neural Bayes estimator targets the posterior median. Alternatively
we could use MSE (targeting the posterior mean). Given that test
evaluation uses a weighted MSE with heavy weight near $\theta = 0$, MAE
may provide more robustness near the boundary.

## Training

```{r nbe-train, eval=FALSE}
estimator <- train(
  estimator,
  sampler   = sampler,
  simulator = simulator,
  m         = 50L,         # each training dataset has k = 50 observations
  epochs    = 100L,
  batchsize = 128L
  # savepath  = "nbe_weights"   # uncomment to save checkpoints
)
```

## Assessing training performance

```{r nbe-risk-plot, eval=FALSE}
# The train() function prints training/validation risk per epoch.
# You can also do a manual risk check on held-out simulated data:

K_val  <- 500
theta_val <- sampler(K_val)
Z_val     <- simulator(theta_val, 50)
theta_hat_val <- estimate(estimator, Z_val)

# Scatter: estimated vs true
plot(as.numeric(theta_val), as.numeric(theta_hat_val),
     pch = 16, cex = 0.5, col = rgb(0, 0, 0, 0.3),
     xlab = expression(theta[true]), ylab = expression(hat(theta)[NBE]),
     main = "NBE: estimated vs. true (validation)")
abline(0, 1, col = "red", lwd = 2)
```

## Apply NBE to test data

```{r nbe-test, eval=FALSE}
# Convert test data to the format expected by NeuralEstimators
# Each element of Z_test should be a 1 x 50 matrix
Z_test_formatted <- lapply(Z_test, function(x) matrix(x, nrow = 1))

theta_nbe <- as.numeric(estimate(estimator, Z_test_formatted))

cat("Summary of NBE estimates:\n")
summary(theta_nbe)
```

# Part 2 – Alternative: Pure-R Neural Network (no Julia)
  
If you cannot install Julia / NeuralEstimators, here is a self-contained
implementation using the **torch** R package (which supports GPU natively).

```{r torch-setup, eval=FALSE}
# install.packages("torch")
library(torch)

# Check GPU
if (cuda_is_available()) {
  device <- torch_device("cuda")
  cat("GPU available! Using CUDA.\n")
} else {
  device <- torch_device("cpu")
  cat("No GPU found. Using CPU.\n")
}
```

```{r torch-deepset, eval=FALSE}
# ---- DeepSets neural network in torch ----
deepset_module <- nn_module(
  "DeepSet",
  initialize = function(d_in = 1, d_out = 1, width = 128) {
    # Inner network (processes each observation)
    self$psi <- nn_sequential(
      nn_linear(d_in, width),
      nn_relu(),
      nn_linear(width, width),
      nn_relu(),
      nn_linear(width, width),
      nn_relu()
    )
    # Outer network (processes aggregated summary)
    self$phi <- nn_sequential(
      nn_linear(width, width),
      nn_relu(),
      nn_linear(width, width),
      nn_relu(),
      nn_linear(width, d_out),
      nn_softplus()        # ensures theta >= 0
    )
  },
  forward = function(x) {
    # x: (batch_size, k, d_in)
    h <- self$psi(x)           # (batch_size, k, width)
    h <- h$mean(dim = 2)       # mean pooling over observations -> (batch_size, width)
    self$phi(h)                # (batch_size, d_out)
  }
)
```

```{r torch-data-generation, eval=FALSE}
# ---- Simulation functions (same logic as before, pure R) ----
rinvgauss_vec <- function(n, mu, lambda) {
  nu <- rnorm(n)
  y  <- nu^2
  x  <- mu + (mu^2 * y) / (2 * lambda) -
    (mu / (2 * lambda)) * sqrt(4 * mu * lambda * y + mu^2 * y^2)
  z  <- runif(n)
  ifelse(z <= mu / (mu + x), x, mu^2 / x)
}

simulate_one_dataset <- function(theta, k = 50) {
  if (theta < 1e-12) {
    z <- rnorm(k)
    1 / (4 * z^2)
  } else {
    rinvgauss_vec(k, mu = 1 / (2 * sqrt(theta)), lambda = 0.5)
  }
}

sample_theta_prior <- function(K) {
  theta <- numeric(K)
  for (i in 1:K) {
    u <- runif(1)
    if (u < 0.10) {
      theta[i] <- 0
    } else if (u < 0.70) {
      theta[i] <- rexp(1, rate = 3)
    } else {
      theta[i] <- runif(1, 0, 2.5)
    }
  }
  theta
}

generate_batch <- function(batch_size, k = 50) {
  thetas <- sample_theta_prior(batch_size)
  data_list <- lapply(thetas, function(th) simulate_one_dataset(th, k))
  X <- do.call(rbind, data_list)   # (batch_size, k)
  list(
    X     = torch_tensor(X, dtype = torch_float())$unsqueeze(3),  # (B, k, 1)
    theta = torch_tensor(matrix(thetas, ncol = 1), dtype = torch_float())  # (B, 1)
  )
}
```

```{r torch-training, eval=FALSE}
# ---- Training loop ----
model     <- deepset_module()$to(device = device)
optimizer <- optim_adam(model$parameters, lr = 1e-3)
scheduler <- lr_step(optimizer, step_size = 30, gamma = 0.5)

n_epochs    <- 100
batch_size  <- 256
steps_per_epoch <- 100
k <- 50

train_losses <- numeric(n_epochs)
val_losses   <- numeric(n_epochs)

# Fixed validation set
val_batch <- generate_batch(2000, k)
val_X     <- val_batch$X$to(device = device)
val_theta <- val_batch$theta$to(device = device)

for (epoch in 1:n_epochs) {
  model$train()
  epoch_loss <- 0
  
  for (step in 1:steps_per_epoch) {
    batch <- generate_batch(batch_size, k)
    X_batch     <- batch$X$to(device = device)
    theta_batch <- batch$theta$to(device = device)
    
    optimizer$zero_grad()
    theta_hat <- model(X_batch)
    
    # Weighted MAE loss (up-weight small theta)
    weights <- 1 / (theta_batch + 0.001)
    loss <- (weights * torch_abs(theta_hat - theta_batch))$mean()
    
    loss$backward()
    optimizer$step()
    
    epoch_loss <- epoch_loss + loss$item()
  }
  
  scheduler$step()
  train_losses[epoch] <- epoch_loss / steps_per_epoch
  
  # Validation
  model$eval()
  with_no_grad({
    val_hat  <- model(val_X)
    val_w    <- 1 / (val_theta + 0.001)
    val_loss <- (val_w * torch_abs(val_hat - val_theta))$mean()
    val_losses[epoch] <- val_loss$item()
  })
  
  if (epoch %% 10 == 0) {
    cat(sprintf("Epoch %3d | Train loss: %.4f | Val loss: %.4f\n",
                epoch, train_losses[epoch], val_losses[epoch]))
  }
}
```

```{r torch-training-plot, eval=FALSE}
plot(1:n_epochs, train_losses, type = "l", col = "blue",
     xlab = "Epoch", ylab = "Weighted MAE Loss",
     main = "Training and Validation Loss")
lines(1:n_epochs, val_losses, col = "red")
legend("topright", legend = c("Train", "Validation"),
       col = c("blue", "red"), lty = 1)
```

```{r torch-apply-test, eval=FALSE}
# ---- Apply trained model to the 1000 test datasets ----
model$eval()

X_test_mat <- do.call(rbind, lapply(Z_test, function(x) as.numeric(x)))
X_test_tensor <- torch_tensor(X_test_mat, dtype = torch_float())$unsqueeze(3)$to(device = device)

with_no_grad({
  theta_nbe <- as.numeric(model(X_test_tensor)$cpu())
})

cat("Summary of NBE estimates:\n")
summary(theta_nbe)
```

# Part 3 – MCMC (10 bonus points)

We implement a Metropolis–Hastings sampler for the posterior of $\theta$
given a dataset. We use the same prior as in the NBE:
$\pi(\theta) \propto 0.10\,\delta_0(\theta) + 0.60\cdot 3e^{-3\theta} + 0.30\cdot\mathbf{1}_{[0,2.5]}(\theta)/2.5$.

Since $\theta$ is continuous and the point mass at 0 complicates mixing,
in practice we use a continuous approximation:
$\pi(\theta) \propto 3e^{-3\theta}\,\mathbf{1}(\theta \ge 0)$
(an Exponential(3) prior), which places heavy mass near 0 and is proper.

## Log-posterior

```{r mcmc-logpost}
log_likelihood <- function(theta, x) {
  k <- length(x)
  if (theta < 0) return(-Inf)
  # Constant terms
  C <- -k/2 * log(4 * pi) - 1.5 * sum(log(x)) - 0.25 * sum(1/x)
  C - theta * sum(x) + k * sqrt(max(theta, 0))
}

log_prior <- function(theta) {
  if (theta < 0) return(-Inf)
  # Exponential(rate=3) prior
  log(3) - 3 * theta
}

log_posterior <- function(theta, x) {
  log_likelihood(theta, x) + log_prior(theta)
}
```

## Metropolis–Hastings sampler

We use a random-walk proposal on the log scale to respect the
non-negativity constraint: propose
$\log\theta^* = \log\theta^{(t)} + \epsilon$, $\epsilon \sim N(0, \sigma^2_\text{prop})$.

```{r mcmc-sampler}
mcmc_sample <- function(x, n_iter = 20000, burn = 5000, 
                         theta_init = NULL, sigma_prop = 0.5) {
  k <- length(x)
  
  # Starting value: MLE
  if (is.null(theta_init)) {
    theta_init <- max(mle_theta(x), 1e-6)
  }
  
  chain <- numeric(n_iter)
  chain[1] <- theta_init
  accept <- 0
  
  for (t in 2:n_iter) {
    theta_current <- chain[t - 1]
    
    # Propose on log scale
    log_theta_star <- log(theta_current) + rnorm(1, 0, sigma_prop)
    theta_star <- exp(log_theta_star)
    
    # Log acceptance ratio (including Jacobian for log-transform)
    log_alpha <- log_posterior(theta_star, x) - log_posterior(theta_current, x) +
      log(theta_star) - log(theta_current)   # Jacobian
    
    if (log(runif(1)) < log_alpha) {
      chain[t] <- theta_star
      accept <- accept + 1
    } else {
      chain[t] <- theta_current
    }
  }
  
  list(chain = chain, accept_rate = accept / (n_iter - 1), burn = burn)
}
```

## Convergence diagnostics for first test dataset

```{r mcmc-diagnostics, fig.width=8, fig.height=8}
# Run MCMC on the first test dataset
res1 <- mcmc_sample(Z_test[[1]], n_iter = 30000, burn = 5000, sigma_prop = 0.3)

cat("Acceptance rate:", round(res1$accept_rate, 3), "\n")

chain_post <- res1$chain[(res1$burn + 1):length(res1$chain)]

par(mfrow = c(2, 2))

# Trace plot
plot(chain_post, type = "l", col = "grey40",
     main = "Trace plot for θ (post burn-in)",
     xlab = "Iteration", ylab = expression(theta))
abline(h = mean(chain_post), col = "red", lwd = 2)

# Histogram
hist(chain_post, breaks = 50, col = "steelblue", border = "white",
     main = "Posterior distribution of θ",
     xlab = expression(theta), probability = TRUE)
abline(v = mean(chain_post), col = "red", lwd = 2)
abline(v = quantile(chain_post, c(0.025, 0.975)), col = "blue", lty = 2)

# ACF
acf(chain_post, lag.max = 100, main = "Autocorrelation")

# Running mean
running_mean <- cumsum(chain_post) / seq_along(chain_post)
plot(running_mean, type = "l", col = "steelblue",
     main = "Running mean",
     xlab = "Iteration", ylab = expression(bar(theta)))

par(mfrow = c(1, 1))

cat("\nPosterior summary for test dataset 1:\n")
cat("  Mean:           ", round(mean(chain_post), 5), "\n")
cat("  Median:         ", round(median(chain_post), 5), "\n")
cat("  95% CI:         [", round(quantile(chain_post, 0.025), 5), ",",
    round(quantile(chain_post, 0.975), 5), "]\n")
cat("  Effective size: ", round(coda::effectiveSize(chain_post), 1), "\n")
```

## Run MCMC for all 1000 test datasets

```{r mcmc-all, eval=FALSE}
# This takes a while — ~10 min depending on chain length
theta_mcmc <- numeric(length(Z_test))

for (i in seq_along(Z_test)) {
  res <- mcmc_sample(Z_test[[i]], n_iter = 20000, burn = 5000, sigma_prop = 0.3)
  chain_post <- res$chain[(res$burn + 1):length(res$chain)]
  theta_mcmc[i] <- mean(chain_post)
  
  if (i %% 100 == 0) cat("Done", i, "of", length(Z_test), "\n")
}

cat("Summary of MCMC posterior means:\n")
summary(theta_mcmc)
```

# Evaluation

Compute the weighted MSE for each method:

$$
\text{MSE} = \frac{1}{1000}\sum_{i=1}^{1000} w(\theta_i)\,(\hat\theta_i - \theta_i)^2, \quad w(\theta_i) = \frac{1}{\theta_i + 0.001}.
$$

```{r evaluation, eval=FALSE}
# You need the true theta values — these come with test_data.RData
# Adjust the object name as needed (e.g., theta_true, true_theta)
# theta_true <- ...

wmse <- function(theta_hat, theta_true, delta = 0.001) {
  w <- 1 / (theta_true + delta)
  mean(w * (theta_hat - theta_true)^2)
}

cat("Weighted MSE — MLE: ", wmse(theta_mle, theta_true), "\n")
cat("Weighted MSE — NBE: ", wmse(theta_nbe, theta_true), "\n")
# cat("Weighted MSE — MCMC:", wmse(theta_mcmc, theta_true), "\n")
```

# Saving estimates for submission

```{r save-results, eval=FALSE}
results <- data.frame(
  dataset = 1:1000,
  MLE     = theta_mle,
  NBE     = theta_nbe
  # MCMC  = theta_mcmc    # uncomment if you ran MCMC
)
write.csv(results, "HW3_P2_estimates.csv", row.names = FALSE)

save(theta_mle, theta_nbe, file = "HW3_P2_estimates.RData")
# save(theta_mle, theta_nbe, theta_mcmc, file = "HW3_P2_estimates.RData")
```

# Uncertainty Comparison (Bonus)

## NBE Uncertainty via Bootstrap

```{r nbe-bootstrap, eval=FALSE}
# Using NeuralEstimators package:
theta_hat_1 <- estimate(estimator, Z_test_formatted[[1]])
boot_mat    <- bootstrap(estimator, Z_test_formatted[[1]], B = 1000)

cat("NBE point estimate for dataset 1:", as.numeric(theta_hat_1), "\n")
cat("Bootstrap 95% CI: [", quantile(boot_mat[1,], 0.025), ",",
    quantile(boot_mat[1,], 0.975), "]\n")
cat("Bootstrap SD:", sd(boot_mat[1,]), "\n")

hist(boot_mat[1,], breaks = 30, col = "steelblue", border = "white",
     main = "NBE Bootstrap Distribution (Dataset 1)",
     xlab = expression(hat(theta)))
abline(v = as.numeric(theta_hat_1), col = "red", lwd = 2)
```

## Comparison Summary

```{r comparison, eval=FALSE}
cat("\n=== Uncertainty Comparison for Test Dataset 1 ===\n\n")

# MCMC results (already computed above)
cat("MCMC:\n")
cat("  Posterior mean:  ", round(mean(chain_post), 5), "\n")
cat("  95% credible CI: [", round(quantile(chain_post, 0.025), 5), ",",
    round(quantile(chain_post, 0.975), 5), "]\n\n")

# NBE results
cat("NBE:\n")
cat("  Point estimate:  ", round(as.numeric(theta_hat_1), 5), "\n")
cat("  95% bootstrap CI:[", round(quantile(boot_mat[1,], 0.025), 5), ",",
    round(quantile(boot_mat[1,], 0.975), 5), "]\n")
```
