---
title: "Homework 3 – Problem 2: Neural Bayes Estimators"
author: "Team Members: Mengyan Jing and Abhisek Banerjee"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(2026)
```

## Problem 1

We have a dataset given to us as peru.txt that consists of variables possibly relating to blood pressures of n = 39 Peruvians who have moved from rural high-altitude areas to urban lower- altitude areas.

Our goal is to construct a coordinate descent algorithm to find parameter estimates for a lasso (L2) regression of systolic blood pressure on the predictors $X_1-X_9$ ($X_1$ and $X_9$ are defined in the question).


### (1) Pseudocode

We have the lasso problem 
$$
\min_{\boldsymbol\beta \in \mathbb{R}^{p}}\;
\tfrac{1}{2}\,\big\|\mathbf{y} - X\boldsymbol\beta\big\|_{2}^{2}
\;+\; \lambda\,\|\boldsymbol\beta\|_{1}.
$$
The non-smooth part $\|\boldsymbol\beta\|_{1} = \sum_{i=1}^{p}|\beta_{i}|$
is separable, so coordinate descent should converge to a
global minimizer. Fixing all $\beta_{j}$, $j\neq i$, the
stationarity condition for $\beta_{i}$ is
$$
0 \;=\; X_{i}^{\top}X_{i}\,\beta_{i} \;+\; X_{i}^{\top}\!\big(X_{-i}\boldsymbol\beta_{-i} - \mathbf{y}\big) \;+\; \lambda\,s_{i},
\qquad s_{i}\in\partial|\beta_{i}|,
$$
which is solved in closed form by soft-thresholding (from the slides of Ryan Tibshirani of Carneige Mellon University):
$$
\beta_{i} \;\leftarrow\;
S_{\lambda/\|X_{i}\|_{2}^{2}}\!\left(
\frac{X_{i}^{\top}\!\big(\mathbf{y}-X_{-i}\boldsymbol\beta_{-i}\big)}
     {X_{i}^{\top}X_{i}}
\right),
\qquad
S_{\tau}(z)=\operatorname{sign}(z)\big(|z|-\tau\big)_{+}.
$$
Maintaining the full residual $\mathbf{r}=\mathbf{y}-X\boldsymbol\beta$,
the update can be rewritten as
$$
\beta_{i} \;\leftarrow\;
S_{\lambda/\|X_{i}\|_{2}^{2}}\!\left(
\frac{X_{i}^{\top}\mathbf{r}}{\|X_{i}\|_{2}^{2}} \;+\; \beta_{i}
\right),
$$
so each coordinate update costs $O(n)$ and one full cycle costs $O(np)$,
matching the cost of one gradient-descent step. Following from the above theory, the pseudo code for our coordinate descent algorithm is as follows:

```text
Inputs: Response variable  y  (in R)
        Design matrix    X  (in R^{n x p}, having columns X_1, ..., X_p)
        Penalty hyperparameter   lambda > 0
        A tolerance limit      (e.g. 1e-7)
        Number of cycles (maximum) K_max

Step 1. We can center y and each column of X (in this may the intercept will drop out)
    so... y    <- y - mean(y)
          X_i  <- X_i - mean(X_i),     i = 1, ..., p

Step 2. Compute column norms from beforehand (which is to be used in every update)
    c_i  <- ||X_i||_2^2 = X_i' X_i,   i = 1, ..., p

Step 3. Now we Initialize everything as zero 
    beta <- (0, 0, ..., 0)            # p-vector of zeros
    r    <- y                         # since beta = 0
    k    <- 0

Step 4. Cyclic coordinate descent
    repeat
        beta_old <- beta
        for i = 1, 2, ..., p do
            # 1. Add coordinate i back into the residual
            r <- r + X_i * beta[i]

            # 2. Univariate OLS coefficient on the partial residual
            z_i <- (X_i' r) / c_i

            # 3. Soft-thresholding update (slide 13)
            beta[i] <- sign(z_i) * max( |z_i| - lambda / c_i , 0 )

            # 4. Subtract the new contribution back out of the residual
            r = K_max

Step 5. Recover the intercept on the original (uncentered) scale
    beta_0 <- mean(y_orig) - sum_i ( beta[i] * mean(X_i_orig) )

Output: (beta_0, beta_1, ..., beta_p)
```













# Background

We consider the exponentially tilted distribution with density:

$$
f(x;\theta) = \sqrt{\frac{1}{4\pi}} \cdot \frac{\exp\!\left(-\frac{1}{4x} - \theta x\right)}{x^{3/2}\,\exp\!\left(-\sqrt{\theta}\right)}, \quad x > 0,\; \theta \ge 0.
$$

The log-density is:

$$
\log f(x;\theta) = -\tfrac{1}{2}\log(4\pi) - \tfrac{3}{2}\log x - \frac{1}{4x} - \theta x + \sqrt{\theta}.
$$

This is an inverse-Gaussian family: for $\theta > 0$,
$X \sim \text{IG}\!\bigl(\mu = \frac{1}{2\sqrt\theta},\; \lambda = \frac{1}{2}\bigr)$.
When $\theta = 0$ it reduces to a Lévy distribution.

Each test dataset has $k = 50$ i.i.d. draws under a common $\theta$.

## Load test data

```{r load-data}
load("test_data.RData")
# Adjust object name if needed:
# Z_test <- test_data
cat("Number of test datasets:", length(Z_test), "\n")
cat("Size of first dataset:", length(Z_test[[1]]), "\n")
head(Z_test[[1]])
```

# Part 1 – Maximum Likelihood Estimation (30 pts)

## Log-likelihood derivation

For $\mathbf{x} = (x_1,\dots,x_k)$:

$$
\ell(\theta) = C(\mathbf{x}) - \theta \sum_i x_i + k\sqrt{\theta},
$$

where $C$ does not depend on $\theta$. Setting $\ell'(\theta) = 0$:

$$
\hat\theta_{\text{MLE}} = \left(\frac{k}{2\sum_i x_i}\right)^2
= \frac{1}{(2\bar{x})^2}.
$$

The second derivative $\ell''(\theta) = -k/(4\theta^{3/2}) < 0$ confirms a maximum.

## Implementation

```{r mle}
mle_theta <- function(x) {
  xbar <- mean(x)
  (1 / (2 * xbar))^2
}

theta_mle <- sapply(Z_test, mle_theta)

cat("Summary of MLE estimates:\n")
summary(theta_mle)
```

```{r mle-hist, fig.width=7, fig.height=4}
hist(theta_mle, breaks = 50, col = "steelblue", border = "white",
     main = "MLE estimates across 1000 test datasets",
     xlab = expression(hat(theta)[MLE]))
```

# Part 2 – Neural Bayes Estimation (45 pts)

## Strategy for minimising weighted MSE

The evaluation metric is:
$$
\text{wMSE} = \frac{1}{1000}\sum_{i=1}^{1000} \frac{(\hat\theta_i - \theta_i)^2}{\theta_i + 0.001}.
$$
At $\theta = 0$ the weight is **1000**; at $\theta = 2$ the weight is **0.5**.
Errors near the boundary dominate the score. Our strategy:

1. **Boundary-heavy training prior** — 20% exact $\theta = 0$, 50% from
   Exp(5) concentrated tightly near 0, 30% uniform for coverage.
2. **Enriched per-observation features** — feed $[x_i,\; 1/x_i,\; \log x_i]$
   instead of just $x_i$. The MLE depends on $1/\bar x$, so giving the network
   direct access to reciprocals helps it learn the sufficient statistic.
3. **Expert summary statistics** — pass MLE $\hat\theta$, $\min(x)$,
   $\text{median}(x)$, and $\text{sd}(\log x)$ as set-level features
   concatenated into the outer network $\phi$. These are highly informative
   for $\theta$ and give the network a strong starting point.
4. **Custom weighted loss** — define a Julia loss that weights squared
   errors by $1/(\theta + 0.001)$, matching the evaluation metric exactly.
5. **Ensemble** — train 3 estimators with different random seeds and
   average their predictions to reduce variance.
6. **Post-processing** — clamp final estimates to $[0, \infty)$.

## Setup

```{r nbe-setup}
library(NeuralEstimators)
# Sys.setenv(JULIA_BINDIR = "/path/to/Julia/bin")
library(JuliaConnectoR)

juliaEval('using NeuralEstimators, Flux')
```

## Simulator

```{r nbe-simulator}
# Inverse-Gaussian sampler (Michael/Schucany/Haas method)
rinvgauss_one <- function(mu, lambda) {
  nu <- rnorm(1)
  y  <- nu^2
  x  <- mu + (mu^2 * y) / (2 * lambda) -
    (mu / (2 * lambda)) * sqrt(4 * mu * lambda * y + mu^2 * y^2)
  if (runif(1) <= mu / (mu + x)) x else mu^2 / x
}

your_sampler <- function(theta) {
  if (theta < 1e-12) {
    z <- rnorm(1)
    return(1 / (4 * z^2))
  } else {
    return(rinvgauss_one(1 / (2 * sqrt(theta)), 0.5))
  }
}
```

## Boundary-heavy training prior

```{r nbe-sampler}
theta_sampler <- function(K) {
  theta <- numeric(K)
  for (i in 1:K) {
    u <- runif(1)
    if (u < 0.20) {
      theta[i] <- 0                        # 20% exact zeros
    } else if (u < 0.70) {
      theta[i] <- rexp(1, rate = 5)        # 50% Exp(5), mean=0.2
    } else {
      theta[i] <- runif(1, 0, 2.5)         # 30% uniform coverage
    }
  }
  matrix(theta, nrow = 1, ncol = K)
}
```

## Enriched features: $[x_i,\; 1/x_i,\; \log x_i]$

Instead of feeding raw $x_i$ (dimension $d=1$), we transform each
observation into a 3-dimensional feature vector. This lets the inner
network $\psi$ immediately see the reciprocal (which drives the MLE)
and the log-scale (which stabilises heavy-tailed inputs from $\theta \approx 0$).

```{r nbe-simulate-enriched}
simulate_enriched <- function(Theta, m) {
  apply(
    Theta, 2,
    function(theta) {
      x <- replicate(m, your_sampler(theta[1]))
      # Stack enriched features: 3 x m matrix
      rbind(x, 1/x, log(x))
    },
    simplify = FALSE
  )
}
```

## Expert summary statistics

We pass set-level summary statistics into the outer network $\phi$.
These are concatenated with the learned neural summary $T(Z)$.

The NeuralEstimators package supports this via the `S` argument in
`DeepSet(psi, phi, S)` where `S` is a Julia function.

```{r nbe-expert-stats}
# R function to compute expert summaries for each dataset
# Returns a vector of expert stats for a single dataset
compute_expert_stats <- function(x_enriched) {
  # x_enriched is a 3 x m matrix: row1 = x, row2 = 1/x, row3 = log(x)
  x <- x_enriched[1, ]

  mle      <- (1 / (2 * mean(x)))^2        # MLE of theta
  min_x    <- min(x)                        # minimum observation
  med_x    <- median(x)                     # median
  sd_logx  <- sd(log(x))                    # SD of log(x)
  inv_mean <- mean(1/x)                     # mean of 1/x
  q10      <- quantile(x, 0.10, names = FALSE)  # 10th percentile

  c(mle, min_x, med_x, sd_logx, inv_mean, q10)
}

# Define expert stats function in Julia that calls R
# NeuralEstimators expects S to take a single data matrix and return a vector
expert_stats_julia <- juliaEval('
  function expert_stats(Z)
    # Z is a d x m matrix (3 x 50 after enrichment)
    x = Z[1, :]  # raw observations in first row

    n    = length(x)
    xbar = sum(x) / n
    mle  = (1 / (2 * xbar))^2

    min_x    = minimum(x)
    med_x    = sort(x)[div(n, 2)]
    log_x    = log.(x)
    sd_logx  = sqrt(sum((log_x .- sum(log_x)/n).^2) / (n - 1))
    inv_mean = sum(1 ./ x) / n
    q10      = sort(x)[max(1, round(Int, 0.10 * n))]

    Float32[mle, min_x, med_x, sd_logx, inv_mean, q10]
  end
')
```

## Custom weighted loss in Julia

We define a loss that directly matches the evaluation metric:
$L(\hat\theta, \theta) = \frac{(\hat\theta - \theta)^2}{\theta + 0.001}$.

```{r nbe-custom-loss}
weighted_loss <- juliaEval('
  function weighted_mse(theta_hat, theta)
    # theta_hat and theta are p x K matrices (1 x K here)
    delta = 0.001f0
    w = 1.0f0 ./ (theta .+ delta)
    mean(w .* (theta_hat .- theta).^2)
  end
')
```

## Neural network architecture with expert stats

```{r nbe-architecture}
d_enriched <- 3   # per-observation features: [x, 1/x, log(x)]
d_expert   <- 6   # expert summary statistics

estimator <- juliaEval(sprintf('
  d  = %d    # enriched observation dimension
  dS = %d    # expert summary stat dimension
  p  = 1     # one parameter: theta
  w  = 128   # hidden layer width

  final_layer = Dense(w, p, softplus)   # enforce theta >= 0

  psi = Chain(
    Dense(d, w, relu),
    Dense(w, w, relu),
    Dense(w, w, relu)
  )

  # phi receives: w (neural summary) + dS (expert stats) = w + dS inputs
  phi = Chain(
    Dense(w + dS, w, relu),
    Dense(w, w, relu),
    Dense(w, w, relu),
    final_layer
  )

  deepset = DeepSet(psi, phi, expert_stats)
  estimator = PointEstimator(deepset)
', d_enriched, d_expert))
```

## Training data

```{r nbe-data}
K <- 20000     # large training set for best performance
m <- 50

theta_train <- theta_sampler(K)
theta_val   <- theta_sampler(K / 5)

Z_train <- simulate_enriched(theta_train, m)
Z_val   <- simulate_enriched(theta_val, m)

cat("Training: K =", ncol(theta_train), ", m =", m, "\n")
cat("Validation: K =", ncol(theta_val), "\n")
```

## Train with custom weighted loss

```{r nbe-train}
estimator <- train(
  estimator,
  theta_train = theta_train,
  theta_val   = theta_val,
  Z_train     = Z_train,
  Z_val       = Z_val,
  loss        = weighted_loss,
  epochs      = 200,
  batchsize   = 64
)
```

## Assess on simulated data

```{r nbe-assess, fig.width=7, fig.height=5}
theta_sim_test <- theta_sampler(1000)
Z_sim_test     <- simulate_enriched(theta_sim_test, m)

assessment <- assess(
  estimator,
  theta_sim_test,
  Z_sim_test,
  estimator_names = "NBE",
  parameter_names = "theta"
)

cat("Simulated test MSE:",
    mean((assessment$estimates$estimate - assessment$estimates$truth)^2), "\n")

# Weighted MSE
est <- assessment$estimates$estimate
tru <- assessment$estimates$truth
wmse_sim <- mean((est - tru)^2 / (tru + 0.001))
cat("Simulated weighted MSE:", wmse_sim, "\n")

plotestimates(
  assessment,
  parameter_labels = c("theta1" = expression(theta))
)
```

## Ensemble: train multiple estimators

We train 3 estimators with different random seeds and training data
realizations, then average their predictions.

```{r nbe-ensemble}
n_ensemble <- 3
estimators <- list()

for (e in 1:n_ensemble) {
  cat("\n=== Training ensemble member", e, "of", n_ensemble, "===\n")

  # Fresh training data each time (simulation on-the-fly effect)
  set.seed(2026 + e * 100)
  th_tr <- theta_sampler(K)
  th_va <- theta_sampler(K / 5)
  Z_tr  <- simulate_enriched(th_tr, m)
  Z_va  <- simulate_enriched(th_va, m)

  # Fresh network
  est_e <- juliaEval(sprintf('
    d  = %d
    dS = %d
    p  = 1
    w  = 128

    psi = Chain(Dense(d, w, relu), Dense(w, w, relu), Dense(w, w, relu))
    phi = Chain(Dense(w + dS, w, relu), Dense(w, w, relu),
                Dense(w, w, relu), Dense(w, p, softplus))

    deepset = DeepSet(psi, phi, expert_stats)
    PointEstimator(deepset)
  ', d_enriched, d_expert))

  est_e <- train(
    est_e,
    theta_train = th_tr,
    theta_val   = th_va,
    Z_train     = Z_tr,
    Z_val       = Z_va,
    loss        = weighted_loss,
    epochs      = 200,
    batchsize   = 64
  )

  estimators[[e]] <- est_e
}
```

## Apply ensemble to test data

```{r nbe-test}
# Enrich test data
Z_test_enriched <- lapply(Z_test, function(x) {
  rbind(as.numeric(x), 1/as.numeric(x), log(as.numeric(x)))
})

# Collect predictions from each ensemble member
preds <- matrix(NA, nrow = n_ensemble, ncol = length(Z_test))
for (e in 1:n_ensemble) {
  preds[e, ] <- as.numeric(estimate(estimators[[e]], Z_test_enriched))
}

# Ensemble average
theta_nbe <- colMeans(preds)

# Post-processing: clamp to [0, Inf)
theta_nbe <- pmax(theta_nbe, 0)

cat("Summary of ensemble NBE estimates:\n")
summary(theta_nbe)

# Individual ensemble member agreement
cat("\nPer-member correlation matrix:\n")
print(cor(t(preds)))
```

```{r nbe-hist, fig.width=7, fig.height=4}
hist(theta_nbe, breaks = 50, col = "coral", border = "white",
     main = "Ensemble NBE estimates across 1000 test datasets",
     xlab = expression(hat(theta)[NBE]))
```

## Fallback: single estimator (if ensemble is too slow)

If you don't want to run the ensemble, use the single estimator from
the first training block:

```{r nbe-single-fallback, eval=FALSE}
Z_test_enriched <- lapply(Z_test, function(x) {
  rbind(as.numeric(x), 1/as.numeric(x), log(as.numeric(x)))
})
theta_nbe <- as.numeric(estimate(estimator, Z_test_enriched))
theta_nbe <- pmax(theta_nbe, 0)
```

# Part 3 – MCMC (10 bonus points)

We use a Metropolis–Hastings sampler with log-scale random walk proposal
and an $\text{Exponential}(3)$ prior.

## Log-posterior

```{r mcmc-logpost}
log_likelihood <- function(theta, x) {
  k <- length(x)
  if (theta < 0) return(-Inf)
  C <- -k/2 * log(4 * pi) - 1.5 * sum(log(x)) - 0.25 * sum(1/x)
  C - theta * sum(x) + k * sqrt(max(theta, 0))
}

log_prior <- function(theta) {
  if (theta < 0) return(-Inf)
  log(3) - 3 * theta          # Exponential(rate=3)
}

log_posterior <- function(theta, x) {
  log_likelihood(theta, x) + log_prior(theta)
}
```

## Metropolis–Hastings sampler

```{r mcmc-sampler}
mcmc_sample <- function(x, n_iter = 20000, burn = 5000,
                         theta_init = NULL, sigma_prop = 0.5) {
  if (is.null(theta_init)) {
    theta_init <- max(mle_theta(x), 1e-6)
  }

  chain <- numeric(n_iter)
  chain[1] <- theta_init
  accept <- 0

  for (t in 2:n_iter) {
    theta_cur <- chain[t - 1]
    log_theta_star <- log(theta_cur) + rnorm(1, 0, sigma_prop)
    theta_star <- exp(log_theta_star)

    log_alpha <- log_posterior(theta_star, x) - log_posterior(theta_cur, x) +
      log(theta_star) - log(theta_cur)

    if (log(runif(1)) < log_alpha) {
      chain[t] <- theta_star
      accept <- accept + 1
    } else {
      chain[t] <- theta_cur
    }
  }

  list(chain = chain, accept_rate = accept / (n_iter - 1), burn = burn)
}
```

## Convergence diagnostics for test dataset 1

```{r mcmc-diagnostics, fig.width=8, fig.height=8}
res1 <- mcmc_sample(Z_test[[1]], n_iter = 30000, burn = 5000, sigma_prop = 0.3)
cat("Acceptance rate:", round(res1$accept_rate, 3), "\n")

chain_post <- res1$chain[(res1$burn + 1):length(res1$chain)]

par(mfrow = c(2, 2))
plot(chain_post, type = "l", col = "grey40",
     main = "Trace plot (post burn-in)",
     xlab = "Iteration", ylab = expression(theta))
abline(h = mean(chain_post), col = "red", lwd = 2)

hist(chain_post, breaks = 50, col = "steelblue", border = "white",
     main = "Posterior of θ", xlab = expression(theta), probability = TRUE)
abline(v = mean(chain_post), col = "red", lwd = 2)
abline(v = quantile(chain_post, c(0.025, 0.975)), col = "blue", lty = 2)

acf(chain_post, lag.max = 100, main = "Autocorrelation")

running_mean <- cumsum(chain_post) / seq_along(chain_post)
plot(running_mean, type = "l", col = "steelblue",
     main = "Running mean", xlab = "Iteration", ylab = expression(bar(theta)))
par(mfrow = c(1, 1))

cat("\nPosterior summary for test dataset 1:\n")
cat("  Mean:  ", round(mean(chain_post), 5), "\n")
cat("  Median:", round(median(chain_post), 5), "\n")
cat("  95% CI: [", round(quantile(chain_post, 0.025), 5), ",",
    round(quantile(chain_post, 0.975), 5), "]\n")
cat("  ESS:   ", round(coda::effectiveSize(chain_post), 1), "\n")
```

## Run MCMC on all 1000 datasets

```{r mcmc-all}
theta_mcmc <- numeric(length(Z_test))

for (i in seq_along(Z_test)) {
  res <- mcmc_sample(Z_test[[i]], n_iter = 20000, burn = 5000, sigma_prop = 0.3)
  chain_post_i <- res$chain[(res$burn + 1):length(res$chain)]
  theta_mcmc[i] <- mean(chain_post_i)
  if (i %% 100 == 0) cat("Done", i, "of", length(Z_test), "\n")
}

cat("Summary of MCMC posterior means:\n")
summary(theta_mcmc)
```

# Uncertainty Comparison (Bonus)

## NBE uncertainty via bootstrap

```{r nbe-bootstrap}
# Use the first ensemble member for bootstrap
theta_hat_1 <- estimate(estimators[[1]], Z_test_enriched[1])
boot_mat    <- bootstrap(estimators[[1]], Z_test_enriched[1], B = 1000)

cat("NBE point estimate:", as.numeric(theta_hat_1), "\n")
cat("Bootstrap 95% CI: [", quantile(boot_mat[1,], 0.025), ",",
    quantile(boot_mat[1,], 0.975), "]\n")
cat("Bootstrap SD:", sd(boot_mat[1,]), "\n")

hist(boot_mat[1,], breaks = 30, col = "coral", border = "white",
     main = "NBE Bootstrap (Dataset 1)", xlab = expression(hat(theta)))
abline(v = as.numeric(theta_hat_1), col = "red", lwd = 2)
```

## Side-by-side comparison

```{r comparison}
cat("\n=== Uncertainty for Test Dataset 1 ===\n\n")
cat("MCMC posterior mean: ", round(mean(chain_post), 5), "\n")
cat("MCMC 95% CI:         [", round(quantile(chain_post, 0.025), 5), ",",
    round(quantile(chain_post, 0.975), 5), "]\n\n")
cat("NBE point estimate:  ", round(as.numeric(theta_hat_1), 5), "\n")
cat("NBE 95% bootstrap CI:[", round(quantile(boot_mat[1,], 0.025), 5), ",",
    round(quantile(boot_mat[1,], 0.975), 5), "]\n")
```

# Evaluation

```{r evaluation}
# theta_true should come with test_data.RData — adjust name as needed
# theta_true <- ...

wmse <- function(theta_hat, theta_true, delta = 0.001) {
  w <- 1 / (theta_true + delta)
  mean(w * (theta_hat - theta_true)^2)
}

# Uncomment once you have theta_true:
# cat("Weighted MSE — MLE:          ", wmse(theta_mle, theta_true), "\n")
# cat("Weighted MSE — NBE (ensemble):", wmse(theta_nbe, theta_true), "\n")
# cat("Weighted MSE — MCMC:         ", wmse(theta_mcmc, theta_true), "\n")
```

# Save estimates for submission

```{r save}
results <- data.frame(
  dataset = 1:1000,
  MLE     = theta_mle,
  NBE     = theta_nbe,
  MCMC    = theta_mcmc
)
write.csv(results, "HW3_P2_estimates.csv", row.names = FALSE)
save(theta_mle, theta_nbe, theta_mcmc, file = "HW3_P2_estimates.RData")
```
